---
title: "Tutorial 12 - Machine Learning"
author: "SFR 605"
---

# Load libraries

```{r, echo=FALSE}
library(ranger)
library(xgboost)
library(vip)
library(pdp)
library(kernelshap)
library(shapviz)
library(tidyverse)
library(lme4)
library(lmerTest)
library(patchwork)
library(here)
```

# Introduction

The aim of this tutorial is to introduce machine learning methods for ecological prediction, with a focus on random forests and model interpretation. We will use soil carbon data from the National Ecological Observatory Network (NEON) to predict soil organic carbon (SOC) concentration from environmental and soil predictors. Along the way we will compare our ML model to the linear mixed models covered earlier in the course and learn to interpret black-box predictions using Shapley values.

`patchwork` lets you combine ggplot objects using `+` for side-by-side layouts and `/` for stacked layouts. We will use it throughout to compare plots. For example:
```
p1 + p2                          ## side by side
p1 / p2                          ## stacked
p1 + p2 + plot_annotation(title = "My combined figure")
```

# Load and explore the data

```{r}
soc_data <- read_csv("data/neon_soil_soc.csv")
glimpse(soc_data)
summary(soc_data)
```

The dataset contains soil physicochemical properties and environmental covariates measured at NEON soil sampling locations across the United States. The variables are:

| Variable | Description | Role |
|---|---|---|
| `soc_pct` | Soil organic carbon (%) | Response |
| `mat_c` | Mean annual temperature (°C) | Predictor |
| `map_mm` | Mean annual precipitation (mm) | Predictor |
| `clay_pct` | Soil clay content (%) | Predictor |
| `sand_pct` | Soil sand content (%) | Predictor |
| `bulk_density` | Soil bulk density (g/cm³) | Predictor |
| `horizon` | Soil horizon (O, A, B) | Grouping |
| `site_id` | NEON site identifier | Grouping |

## Missing values

```{r}
soc_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  filter(n_missing > 0)
```

For now we will use complete cases only, removing any observation with a missing value in any predictor.

```{r}
soc_clean <- soc_data %>% drop_na()
cat("Observations after removing NAs:", nrow(soc_clean), "\n")
cat("Observations removed:", nrow(soc_data) - nrow(soc_clean), "\n")
```

### Task 1
How many observations have missing values? Which variables are affected? What assumption does listwise deletion make about the mechanism of missingness?

## Response distribution

```{r}
ggplot(soc_clean, aes(x = soc_pct)) +
  geom_histogram(bins = 30, fill = "#1D9E75", color = "white") +
  labs(x = "Soil organic carbon (%)", y = "Count",
       title = "Distribution of the response variable") +
  theme_minimal()
```

### Task 2
Describe the distribution of `soc_pct`. Is it symmetric? Random forests are generally robust to skewed responses — note any concerns but proceed without transformation for now.

## Predictor relationships

```{r}
soc_clean %>%
  select(soc_pct, mat_c, map_mm, clay_pct, sand_pct, bulk_density) %>%
  pairs(pch = 16, cex = 0.4, col = alpha("#378ADD", 0.5),
        main = "Pairwise relationships among predictors and response")
```

### Task 3
From the pairs plot, identify: (a) one predictor that appears to have a strong relationship with `soc_pct`, and (b) two predictors that are correlated with each other. Note: correlated predictors are handled differently by random forests than by linear models. We will return to this when we discuss variable importance.

# Fitting a Random Forest

## Train/test split

Before fitting any model, we set aside 20% of the data as a held-out test set. This test set will not be touched until the evaluation section.

```{r}
predictors <- c("mat_c", "map_mm", "clay_pct", "sand_pct", "bulk_density")
response   <- "soc_pct"

set.seed(605)
n      <- nrow(soc_clean)
test_i <- sample(seq_len(n), size = floor(0.2 * n))
train  <- soc_clean[-test_i, ]
test   <- soc_clean[ test_i, ]

cat("Training observations:", nrow(train), "\n")
cat("Test observations:    ", nrow(test),  "\n")
```

## Fit the model

`ranger` is a fast C++ implementation of random forests. The key arguments are `num.trees` (number of trees; 500–1000 is usually sufficient), `mtry` (number of predictors randomly sampled at each split; default is p/3 for regression), `min.node.size` (minimum observations in a terminal node, controlling tree depth), `importance` (method for variable importance), and `keep.inbag = TRUE` which is required later for SHAP computation.

```{r}
rf_fit <- ranger(
  formula       = soc_pct ~ mat_c + map_mm + clay_pct + sand_pct + bulk_density,
  data          = train,
  num.trees     = 500,
  mtry          = 3,
  min.node.size = 5,
  importance    = "permutation",
  keep.inbag    = TRUE,
  seed          = 605
)
rf_fit
```

### Task 4
Examine the `ranger` output. (a) What is the OOB prediction error (reported as MSE)? Convert it to RMSE and compare it to the standard deviation of `soc_pct` in the training set. A useful heuristic: RMSE / SD < 0.5 suggests reasonable predictive skill. (b) What is the OOB R²? How does "out-of-bag" differ from a held-out test set as a means of assessing model performance?

## Hyperparameter tuning

The default `mtry = 3` may not be optimal for this dataset. The code below performs a simple grid search over all possible `mtry` values, using OOB error to evaluate each.

```{r}
mtry_vals  <- 1:length(predictors)
oob_errors <- numeric(length(mtry_vals))

for (i in seq_along(mtry_vals)) {
  fit_i <- ranger(
    formula    = soc_pct ~ mat_c + map_mm + clay_pct + sand_pct + bulk_density,
    data       = train,
    num.trees  = 500,
    mtry       = mtry_vals[i],
    keep.inbag = TRUE,
    seed       = 605
  )
  oob_errors[i] <- fit_i$prediction.error
}

data.frame(mtry = mtry_vals, oob_rmse = sqrt(oob_errors)) %>%
  ggplot(aes(x = mtry, y = oob_rmse)) +
  geom_line(color = "#378ADD", linewidth = 1) +
  geom_point(size = 2.5, color = "#378ADD") +
  geom_vline(xintercept = mtry_vals[which.min(oob_errors)],
             linetype = "dashed", color = "#D85A30") +
  labs(x = "mtry (predictors sampled per split)", y = "OOB RMSE",
       title = "Hyperparameter tuning: effect of mtry on OOB error") +
  theme_minimal()
```

### Task 5
(a) What is the optimal `mtry` based on OOB error? (b) Refit the random forest using the optimal `mtry`, storing it as `rf_fit_tuned`. Evaluate it on the test set and compare test R² and RMSE to the default model. How much does performance change? (c) Why is it acceptable to use OOB error for tuning `mtry` rather than needing a separate validation set?

# Variable Importance

## Permutation importance

Permutation importance measures how much prediction error increases when each predictor is randomly shuffled, breaking its relationship with the response. A large increase in error indicates that the predictor is important.

```{r}
vip(rf_fit,
    num_features = length(predictors),
    geom         = "col",
    aesthetics   = list(fill = "#378ADD", color = "white")) +
  labs(title = "Permutation variable importance",
       x = "Predictor", y = "Importance (increase in MSE when shuffled)") +
  theme_minimal()
```

### Task 6
Interpret the variable importance plot. (a) Which predictor has the highest importance? Does this match your expectation from the pairs plot (Task 3)? (b) `bulk_density` dominates both importance metrics. Is this ecologically meaningful, or does it suggest a potential problem? Consider: bulk density is known to be strongly negatively correlated with SOC (high organic matter soils tend to have low bulk density). Does this make bulk density a true *driver* of SOC, or more of a co-measurement of the same underlying property? Would you include it in a predictive model?

## Impurity-based importance

Permutation importance is generally the more reliable of the two importance methods. We can compare it to impurity-based importance (which measures how often each variable is used for splits, weighted by improvement) to see whether they agree.

```{r}
rf_impurity <- ranger(
  formula    = soc_pct ~ mat_c + map_mm + clay_pct + sand_pct + bulk_density,
  data       = train,
  num.trees  = 500,
  mtry       = 3,
  importance = "impurity",
  seed       = 605
)

imp_perm     <- rf_fit$variable.importance
imp_impurity <- rf_impurity$variable.importance

data.frame(
  predictor   = names(imp_perm),
  permutation = imp_perm / max(imp_perm),
  impurity    = imp_impurity / max(imp_impurity)
) %>%
  pivot_longer(-predictor, names_to = "method", values_to = "importance") %>%
  mutate(predictor = fct_reorder(predictor, importance, .fun = max)) %>%
  ggplot(aes(x = importance, y = predictor, fill = method)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(permutation = "#378ADD", impurity = "#D85A30")) +
  labs(title = "Permutation vs. impurity-based importance (both normalized 0–1)",
       x = "Importance (normalized)", y = NULL, fill = "Method") +
  theme_minimal()
```

### Task 7
Do the two importance methods agree? If they disagree for any predictor, which method do you trust more and why? Note: impurity-based importance is known to be biased toward high-cardinality continuous variables (i.e., variables with a high percentage of unique values). Does that explain any discrepancy you see?

# Partial Dependence Plots

Variable importance tells us *that* a predictor matters, but not *how*. Partial dependence plots (PDPs) show the marginal effect of one predictor on the response, averaged over all other predictors. Note that `pdp::partial()` must be called with the package prefix since `purrr` (loaded as part of `tidyverse`) also has a `partial()` function.

```{r}
pdp_top <- pdp::partial(
  rf_fit,
  pred.var        = "bulk_density",
  train           = train,
  grid.resolution = 30
)

autoplot(pdp_top, smooth = TRUE) +
  labs(x = "Bulk density (g/cm³)", y = "Predicted SOC (%)",
       title = "Partial dependence: bulk density vs. SOC") +
  theme_minimal()
```

### Task 8
Using the code above as a template, make PDP plots for two more predictors of your choice. For each of the three PDPs (including bulk density): (a) describe the shape of the relationship (linear, unimodal, threshold, etc.), and (b) describe where uncertainty is widest and why that might be.

## Two-way interaction PDP

PDPs can also visualize interactions between two predictors simultaneously. Two-predictor PDPs must be plotted with `ggplot` using `geom_tile()` rather than `autoplot()`.

```{r}
pdp_2d <- pdp::partial(
  rf_fit,
  pred.var        = c("mat_c", "map_mm"),
  train           = train,
  grid.resolution = 20
)

ggplot(pdp_2d, aes(x = mat_c, y = map_mm, fill = yhat)) +
  geom_tile() +
  scale_fill_gradient2(low = "#378ADD", mid = "white", high = "#D85A30",
                       midpoint = median(pdp_2d$yhat)) +
  labs(x = "Mean annual temperature (°C)", y = "Mean annual precipitation (mm)",
       fill = "Predicted\nSOC (%)", title = "Two-way partial dependence: MAT × MAP") +
  theme_minimal()
```

### Task 9
Using the two-way PDP above as a template, generate a two-way PDP for `bulk_density` and `mat_c`. Does the effect of bulk density on SOC appear to depend on temperature, or is it consistent across the temperature gradient?

# Evaluating on the Test Set

Now we use the held-out test set for an honest estimate of generalization performance. We save the predicted vs. observed plot as `p_rf_obs` so we can reuse it alongside the LMM plot in the next section.

```{r}
test_preds <- predict(rf_fit, data = test)$predictions

rmse <- sqrt(mean((test_preds - test$soc_pct)^2))
mae  <- mean(abs(test_preds - test$soc_pct))
r2   <- cor(test_preds, test$soc_pct)^2

cat("Test set performance\n")
cat("--------------------\n")
cat(sprintf("RMSE: %.3f\n", rmse))
cat(sprintf("MAE:  %.3f\n", mae))
cat(sprintf("R²:   %.3f\n", r2))

p_rf_obs <- data.frame(observed = test$soc_pct, predicted = test_preds) %>%
  ggplot(aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.6, color = "#378ADD") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Observed SOC (%)", y = "Predicted SOC (%)",
       title = "Random forest",
       subtitle = sprintf("R² = %.3f | RMSE = %.3f", r2, rmse)) +
  theme_minimal()

p_rf_obs
```

### Task 10
(a) Compare the test set R² to the OOB R² from the model fit above. Are they similar? If the test R² is substantially lower, what might explain the gap? (b) Are there systematic patterns in the residuals, i.e., does the model consistently over- or under-predict at certain SOC values?

# Comparison to a Linear Mixed Model

Comparing an ML model to a simpler baseline is one of the most informative things we can do. Here we fit a linear mixed model with `site_id` as a random effect, the same structure you explored earlier in the course. Because our predictors are on very different scales (`map_mm` ranges from ~100 to 2400 while `bulk_density` ranges from ~0 to 2), we rescale before fitting. Note that we did not need to do this for the random because random forests are scale-invariant (they only use the rank order of splits). Linear models are not, because coefficients depend directly on the units of the predictors.

```{r}
train_scaled <- train %>% mutate(across(all_of(predictors), scale))
test_scaled  <- test  %>% mutate(across(all_of(predictors), scale))

lmm_fit <- lmer(
  soc_pct ~ mat_c + map_mm + clay_pct + sand_pct + bulk_density + (1 | site_id),
  data = train_scaled,
  REML = FALSE
)
summary(lmm_fit)

lmm_preds <- predict(lmm_fit, newdata = test_scaled, allow.new.levels = TRUE)
lmm_rmse  <- sqrt(mean((lmm_preds - test$soc_pct)^2))
lmm_r2    <- cor(lmm_preds, test$soc_pct)^2

cat("Model comparison (test set)\n")
cat("---------------------------\n")
cat(sprintf("Random forest — RMSE: %.3f | R²: %.3f\n", rmse, r2))
cat(sprintf("Linear LMM    — RMSE: %.3f | R²: %.3f\n", lmm_rmse, lmm_r2))

p_lmm_obs <- data.frame(observed = test$soc_pct, predicted = lmm_preds) %>%
  ggplot(aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.6, color = "#D85A30") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Observed SOC (%)", y = "Predicted SOC (%)",
       title = "Linear mixed model",
       subtitle = sprintf("R² = %.3f | RMSE = %.3f", lmm_r2, lmm_rmse)) +
  theme_minimal()

p_rf_obs + p_lmm_obs +
  plot_annotation(title = "Predicted vs. observed: RF vs. LMM")
```

### Task 11
(a) What is the test set performance of each model? (b) Do LMM assumptions appear to be met based on your exploration of the data above? (c) Under what circumstances might you choose the RF over the LMM, or vice versa? Note: the regularization that keeps the LMM from overfitting (REML shrinkage on random effects) is analogous to the regularization penalty in LASSO/ridge regression. The random forest achieves implicit regularization through bagging and random feature selection.

# SHAP Values

## Global vs. local interpretation

Before computing SHAP values, it is worth being precise about what question we are asking.

| Level | Question | Method |
|---|---|---|
| **Global** | Which predictors matter most, on average? | Variable importance, PDPs |
| **Global** | What is the average effect of predictor X? | Partial dependence plot |
| **Local** | Why did the model make *this* prediction? | SHAP waterfall plot |
| **Local → Global** | How does X affect predictions across all observations? | SHAP beeswarm / dependence plot |

## Fit a gradient boosted model

We also fit an `xgboost` model on the same training data so we can compare SHAP values across two different algorithms and test whether our ecological conclusions depend on algorithm choice.

```{r}
X_train <- train %>% select(all_of(predictors)) %>% as.matrix()
y_train <- train[[response]]
X_test  <- test  %>% select(all_of(predictors)) %>% as.matrix()

xgb_fit <- xgboost(
  x                = X_train,
  y                = y_train,
  nrounds          = 300,
  max_depth        = 4,
  learning_rate    = 0.05,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  objective        = "reg:squarederror"
)
```

## Compute SHAP values

`shapviz` provides a clean interface for visualizing SHAP values. For both `ranger` and `xgboost` models, SHAP values are computed via `kernelshap`, which uses a model-agnostic sampling approach to approximate the contribution of each predictor to each prediction. Because it is an approximation, results have some variance - we will explore this later. The `bg_X` argument specifies the background dataset used as a reference for marginalizing over predictors; we always use the training data for this, even when computing SHAP values for test set observations.

```{r}
ks_rf    <- kernelshap(rf_fit,  X = train[, predictors], bg_X = train[, predictors])
shap_rf  <- shapviz(ks_rf)

ks_xgb   <- kernelshap(xgb_fit, X = X_train, bg_X = X_train)
shap_xgb <- shapviz(ks_xgb)
```

The efficiency property of Shapley values guarantees that the SHAP values for all predictors sum exactly to (prediction − baseline) for every observation. The code below verifies this holds for our model: a maximum discrepancy of zero confirms the SHAP values are internally consistent and that credit for each prediction has been fully and fairly distributed across predictors.

```{r}
shap_sum <- rowSums(shap_rf$S) + shap_rf$baseline
rf_preds <- predict(rf_fit, data = train)$predictions
cat(sprintf("Max discrepancy between SHAP sum and prediction: %.6f\n",
            max(abs(shap_sum - rf_preds))))
```

## Global SHAP importance

```{r}
sv_importance(shap_rf, kind = "bar", fill = "#378ADD") +
  labs(title = "Mean |SHAP| — global feature importance (random forest)") +
  theme_minimal()
```

### Task 12
Compare this SHAP importance ranking to the permutation importance ranking from Task 6. Do they agree? If any predictors change rank substantially, note which ones and suggest why. Think about how correlated predictors are handled differently by permutation importance vs. SHAP.

## SHAP beeswarm plot

The beeswarm plot shows both importance and the direction of each predictor's effect. Each dot is one observation; position on the x-axis is the SHAP value; color encodes the predictor's actual value (red = high, blue = low). Dots to the right indicate that variable increased the prediction for that observation; dots to the left indicate it decreased the prediction. Red dots on the right indicate a positive relationship; blue dots on the right indicate a negative relationship. Wide horizontal spread means that predictor's effect varies a lot across observations.

```{r}
sv_importance(shap_rf, kind = "beeswarm", alpha = 0.5) +
  labs(title = "SHAP beeswarm plot — random forest") +
  theme_minimal()
```

### Task 13
Using `sv_importance()` with `kind = "beeswarm"`, generate the equivalent beeswarm plot for the `xgboost` model (`shap_xgb` is already computed above). Use `patchwork` to place both plots side by side, then describe any differences in the direction or magnitude of effects between the two models.

# Local Interpretation

The beeswarm plot gives us a global picture. The **waterfall plot** explains a single prediction — why did the model predict what it did for this specific observation?

```{r}
ks_rf_test   <- kernelshap(rf_fit, X = test[, predictors], bg_X = train[, predictors])
shap_rf_test <- shapviz(ks_rf_test)

test_preds2 <- predict(rf_fit, data = test)$predictions
high_soc_i  <- which.max(test_preds2)
low_soc_i   <- which.min(test_preds2)

sv_waterfall(shap_rf_test, row_id = high_soc_i) +
  labs(title = "Why did the model predict HIGH SOC here?") +
  theme_minimal()

sv_waterfall(shap_rf_test, row_id = low_soc_i) +
  labs(title = "Why did the model predict LOW SOC here?") +
  theme_minimal()

cat("High SOC observation predictor values:\n")
print(test[high_soc_i, predictors])
cat("\nLow SOC observation predictor values:\n")
print(test[low_soc_i, predictors])
```

### Task 14
(a) For each waterfall plot, which predictors contributed to pushing the prediction above the baseline vs. below? (b) Choose two intermediate SOC points from the dataset and make waterfall plots for them. How do the predictor contributions compare with the high and low SOC waterfall plots?

## Force plot

The force plot conveys the same information as the waterfall plot in a more compact form.

```{r}
sv_force(shap_rf_test, row_id = high_soc_i) +
  labs(title = "Force plot — high SOC observation") +
  theme_minimal()
```

### Task 15
Generate the force plot for the low-SOC observation. Which format (waterfall or force) do you find more intuitive for communicating results to a non-technical audience such as a land manager? Which would you include in a scientific paper?

# SHAP Dependence Plots vs. Partial Dependence Plots

SHAP dependence plots and PDPs both show how a predictor relates to the response, but they answer slightly different questions and can diverge when predictors are correlated. A SHAP dependence plot shows the SHAP value for one predictor (y-axis) vs. the predictor's actual value (x-axis), with points colored by a second predictor to reveal interactions. Below we plot SHAP dependence for MAT and MAP, then compare the MAT SHAP dependence plot to the MAT PDP side by side.

```{r}
sv_dependence(shap_rf, v = "mat_c", color_var = "map_mm") +
  labs(x = "Mean annual temperature (°C)", y = "SHAP value for MAT",
       title = "SHAP dependence: MAT (colored by MAP)") +
  theme_minimal()

sv_dependence(shap_rf, v = "map_mm", color_var = "mat_c") +
  labs(x = "Mean annual precipitation (mm)", y = "SHAP value for MAP",
       title = "SHAP dependence: MAP (colored by MAT)") +
  theme_minimal()

pdp_mat <- pdp::partial(rf_fit, pred.var = "mat_c", train = train,
                         grid.resolution = 40)

p_pdp <- autoplot(pdp_mat, smooth = TRUE) +
  labs(x = "Mean annual temperature (°C)", y = "Predicted SOC (%)",
       title = "Partial dependence plot") +
  theme_minimal()

p_shap <- sv_dependence(shap_rf, v = "mat_c", color_var = NULL) +
  labs(x = "Mean annual temperature (°C)", y = "SHAP value for MAT",
       title = "SHAP dependence plot") +
  theme_minimal()

p_pdp + p_shap +
  plot_annotation(title = "PDP vs. SHAP dependence for mean annual temperature")
```

### Task 16
Following the example above for `mat_c`, generate a side-by-side PDP and SHAP dependence plot for `bulk_density`. (a) Do the two methods show the same general shape? (b) Color the SHAP dependence plot by `mat_c`. Does the effect of bulk density appear to vary with temperature? Describe any interaction pattern you see and suggest an ecological mechanism.

# Cross-Model Comparison

SHAP is model-agnostic. Both `shap_rf` and `shap_xgb` are already computed. Here we compare mean absolute SHAP values across both models to test whether our ecological interpretation depends on algorithm choice. Findings that are robust to algorithm choice are generally considered more trustworthy. This property is sometimes called **algorithmic stability**.

```{r}
imp_rf  <- colMeans(abs(shap_rf$S))
imp_xgb <- colMeans(abs(shap_xgb$S))

data.frame(
  predictor = predictors,
  rf        = imp_rf[predictors],
  xgb       = imp_xgb[predictors]
) %>%
  pivot_longer(-predictor, names_to = "model", values_to = "mean_abs_shap") %>%
  mutate(predictor = fct_reorder(predictor, mean_abs_shap, .fun = max),
         model     = recode(model, rf = "Random forest", xgb = "Gradient boosting")) %>%
  ggplot(aes(x = mean_abs_shap, y = predictor, fill = model)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Random forest" = "#378ADD",
                               "Gradient boosting" = "#1D9E75")) +
  labs(x = "Mean |SHAP value|", y = NULL, fill = "Model",
       title = "SHAP importance: random forest vs. gradient boosting") +
  theme_minimal()
```

### Task 17
Do the two models agree on which predictors are most important? Identify any predictors ranked substantially differently across models. When would you be *more* concerned about algorithmic instability?

# Limitations and Responsible Use of SHAP

Because `kernelshap` is an approximation, SHAP values have some variance across runs. The code below demonstrates this by refitting the random forest with a different random seed and comparing SHAP values for a single observation.

```{r}
rf_fit2 <- ranger(
  formula       = soc_pct ~ mat_c + map_mm + clay_pct + sand_pct + bulk_density,
  data          = train,
  num.trees     = 500,
  mtry          = 3,
  min.node.size = 5,
  keep.inbag    = TRUE,
  seed          = 999
)

ks_rf2   <- kernelshap(rf_fit2, X = train[, predictors], bg_X = train[, predictors])
shap_rf2 <- shapviz(ks_rf2)

data.frame(
  predictor    = predictors,
  shap_seed605 = shap_rf$S[1, predictors],
  shap_seed999 = shap_rf2$S[1, predictors]
) %>%
  mutate(difference = shap_seed605 - shap_seed999) %>%
  arrange(desc(abs(difference)))
```

### Task 18
The table above compares SHAP values for a single observation across two random forest fits that differ only in their random seed. (a) How large are the differences? Are they ecologically meaningful? (b) Implement a robustness check: refit the random forest three more times with seeds 111, 222, and 333. For each fit, compute SHAP values on the training set and extract the SHAP value for `bulk_density` for observation 1. Report the mean and standard deviation across all five fits (including seeds 605 and 999). Are there any conclusions you can be confident about?