---
title: "SFR Preliminary Analysis"
author: "Sander Elliott"
date: "2026-04-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r}
library(tidyr)
library(dplyr)
library(glatos)
library(sf)
library(tidyverse)
library(dataRetrieval)
library(rosm)
library(raster) 
library(brms)
library(patchwork)
```

# Data Prep

## Raw Data

```{r}
SNShst <- read.csv("data/SNShst_allyrs.csv")
SNSfsh <- read.csv("data/SNSfsh_allyrs.csv")
glimpse(SNShst) ## Detection history
glimpse(SNSfsh) ## Fish information
```

## Detection History

```{r}
## Remove columns for types of tags not used in analysis
## Only keep telemetry detections
SNSdetcln <- SNShst %>% 
  filter(Event == "Detection") %>% 
  dplyr::select(-Event, -Species, -tagtype, -SensorType, -avgSensorValue, -avgComputedValue, -Frequency, -avgPower,
         -LocationCode, -AntennaID, -AltFishID)

```

### Prep Glatos - Convert coordiantes and time

```{r}
detglat_sf <- SNSdetcln %>% 
  mutate(
    detection_timestamp_utc = as.POSIXct(LastTS),
    transmitter_id = as.character(IDCode),
    transmitter_codespace = as.character(TagId),
    receiver_sn = RxID,
    animal_id = FishID
  ) %>% 
  filter(!is.na(Easting)) %>% 
  st_as_sf(coords = c("Easting", "Northing"), crs = 26919) %>%   # UTM 19N
  st_transform(4326) %>%                                         # WGS84 lat/long
  mutate(
    deploy_long = st_coordinates(.)[,1],
    deploy_lat  = st_coordinates(.)[,2]
  ) %>% 
  filter(deploy_lat < 45.1) ## some false detections way north

glat_sf <- detglat_sf %>% 
  mutate(array = case_when(deploy_lat < 44.49 & deploy_long > -69.2 ~ 
                             "penob_bay",
                           deploy_lat > 44.49 & deploy_long > -69.2 ~
                             "penob_riv",
                           deploy_long < -69.2 ~
                             "ken_r"))


detglat_sf_3857a <- st_transform(glat_sf, 3857)

glimpse(glat_sf)
```

### Use GLATOS

```{r}
glat_events <- detection_events(glat_sf, location_col = "array", condense = TRUE) %>% 
  mutate(m_leave = month(last_detection),
         y_leave = year(last_detection), 
         d_leave = yday(last_detection),
         m_leave = month(last_detection),
         y_arrive = year(first_detection),
         d_arrive = yday(first_detection))
glimpse(glat_events)
```

### Plot Arrays

```{r}
bb <- st_bbox(detglat_sf)
osm <- osm.raster(bb, type = "cartolight")
osm_df <- as.data.frame(osm, xy = TRUE)

d_map <- ggplot() +
  geom_raster(data = osm_df,
              aes(x = x, y = y,
                  fill = rgb(layer.1/255, layer.2/255, layer.3/255))) +
  scale_fill_identity() +
  geom_sf(data = detglat_sf_3857a, aes(color = array)) +
  scale_color_manual(
    name = "System",
    values = c(
      "ken_r"     = 3,
      "penob_bay" = 4,
      "penob_riv" = 2
    ),
    labels = c(
      "ken_r"     = "Kennebec River",
      "penob_bay" = "Penobscot Bay",
      "penob_riv" = "Penobscot River"
    )
  ) +
  scale_x_continuous(n.breaks = 4) +
  coord_sf(crs = 3857) +
  theme_classic() +
  labs(
    title = "Detection Locations",
    x = "Degrees Longitude",
    y = "Degrees Latitude"
  ) +
  theme(legend.position = "none")

d_map
```

### Get Penobscot Departures

```{r}
penob_events <- glat_events %>%
  arrange(animal_id, last_detection) %>% 
  group_by(animal_id) %>%
  mutate(is_last = row_number() == n()) %>%   # TRUE only for final row
  filter(!(is_last & location == "penob_riv")) %>%  # drop only final penob_riv
  dplyr::select(-is_last) %>%
  ungroup() %>% 
  filter(location == "penob_riv")

```

### Plot Events

```{r}
penob_plot <- glat_events %>% 
    arrange(animal_id, last_detection) %>% 
  group_by(animal_id) %>%
  mutate(is_last = row_number() == n()) %>%   # TRUE only for final row
  filter(!(is_last & location == "penob_riv")) %>%  # drop only final penob_riv
  dplyr::select(-is_last) %>%
  ungroup() %>% 
  filter(animal_id %in% penob_events$animal_id)

d_time <- ggplot(penob_plot, aes(
  y = animal_id,
  x = first_detection,
  xend = last_detection,
  yend = animal_id,
  color = location)) +
  geom_segment(linewidth = 1.2, alpha = 0.8) +
  geom_point(aes(x = first_detection), size = 2) + 
  geom_point(aes(x = last_detection), size = 2) +
  labs(
    title = "Detection Timeline",
    x = "Detection Interval",
    y = "Individual",
    color = "System"
  ) +
  scale_color_manual(
    values = c(
      "ken_r"     = 3,
      "penob_bay" = 4,
      "penob_riv" = 2
    ),
    labels = c(
      "ken_r"     = "Kennebec River",
      "penob_bay" = "Penobscot Bay",
      "penob_riv" = "Penobscot River"
    )
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_blank(),
    legend.position = "none",
    aspect.ratio = 0.5
  )

d_time
```

#### Figure 1

```{r}
png("outputs/map_time.png", width = 2700, height = 1000, res = 300)

(d_map | d_time) +
  theme(legend.position = "right")
dev.off()
```


## Fish

### Keep important data

```{r}
SNSfshcln <- SNSfsh %>% 
  dplyr::select(FishID, TagID, CaptureDate, ForkLength, TotalLength, Mass, Recapture, InitialRelease, Sex)

```

### Prep fish for Join

```{r}
departfsh <- SNSfshcln %>% 
  filter(FishID %in% penob_events$animal_id & InitialRelease == 1)%>%
  rename(animal_id = FishID)
```

## Seasons

### Join fish and detections

```{r warning=FALSE}
depart_full <- penob_events %>%
  left_join(departfsh, by = "animal_id") %>% 
  mutate(CaptureDate = as.POSIXct(CaptureDate, format = "%Y-%m-%d %H:%M:%S"),
         day_cap = as.numeric(difftime(last_detection, CaptureDate, units = "days")),
         day_syst = res_time_sec/86400) %>% 
  filter(day_syst > 2)
glimpse(depart_full)
```

### Create Seasons

```{r}
depart_ssn <- depart_full %>% 
  mutate(season = case_when(d_leave > 240 ~ "fall",
                            d_leave < 136 ~ "spring",
                            d_leave > 136 & d_leave < 240 ~ "summer"),
         arrive_ssn = case_when(d_arrive >= 240 ~ "fall",
                                d_arrive <= 136 ~ "spring",
                                d_arrive > 136 & d_arrive < 240 ~ "summer")) %>% 
  rename(year = y_leave)
glimpse(depart_ssn)
```

## Flow

### Raw USGS stream gauge

```{r}
API_USGS_PAT = "UMOm8b4Xd4NQmdHXvgJgloj4ROEJXneryqIYXVhU"


site <- "01036390"
pcode <- "00065"  # gage height

raw <- readNWISdv("01036390", "00065", "1900-01-01", Sys.Date())

```

### Add seasons

```{r}
dat <- raw %>%
  rename(gage_height = X_00065_00003) %>% 
  mutate(
    jday = yday(Date),
    year = year(Date),
    season = case_when(
      jday >= 87  & jday <= 136 ~ "spring",
      jday >= 137 & jday <= 217 ~ "summer",
      jday >= 254 & jday <= 321 ~ "fall",
      TRUE ~ NA_character_)
  )

season_dat <- dat %>% filter(!is.na(season))

seasonal_means <- season_dat %>%
  group_by(year, season) %>%
  summarize(
    seasonal_mean = mean(gage_height, na.rm = TRUE),
    n_days = n(),
    .groups = "drop"
  )
  
```

#### Figure 2

```{r}
seasonal_plot <- seasonal_means %>% 
  mutate(season = factor(season, levels = c("spring", "summer", "fall")))

png("outputs/usgs_flow.png", width = 2000, height = 1500, res = 300)
  
ggplot(seasonal_plot, aes(x = year, y = seasonal_mean, color = season)) +
  geom_line() +
  theme_classic() +
  scale_color_manual(
    values = c(
      "fall"  = 5,
      "spring" = 6,
      "summer" = 7
    ),
    labels = c(
      "fall"     = "Fall",
      "spring" = "Spring",
      "summer"   = "Summer"
    )) +
  labs(title = "Seasonal Mean Gage Height\nUSGS 01036390 Penobscot River at Eddington, ME",
       y = "Gage Height (ft)",
       x = "Year",
       color = "Season")
dev.off()
```


### Scale flow seasons

```{r}

flow_wide <- seasonal_means %>%
  dplyr::select(-n_days) %>% 
  tidyr::pivot_wider(
    names_from = season,
    values_from = seasonal_mean,
    names_glue = "{season}_flow"
  ) %>% 
  mutate(
    spring_flow_z = as.numeric(scale(spring_flow)),
    summer_flow_z = as.numeric(scale(summer_flow)),
    fall_flow_z   = as.numeric(scale(fall_flow))
  )
glimpse(flow_wide)

depart_flow <- depart_ssn %>%
  left_join(flow_wide %>% 
              dplyr::select(year, spring_flow_z, summer_flow_z, fall_flow_z), by = "year")

glimpse(depart_flow)

```

### Figure 3

```{r warning=FALSE}
d_hist <- ggplot(depart_flow, aes(x = d_leave)) +
  geom_histogram(aes(y = ..density..), bins = 52, fill = "grey80", color = "white") +
  geom_density(color = "blue", size = 1) +
  coord_cartesian(xlim = c(1, 365)) +
  geom_vline(xintercept =240) +
  geom_vline(xintercept =136) +
  theme_classic() +
  labs(
    title = "Departure Timing Distribution",
    subtitle = "Shortnose Sturgeon from Penobscot River",
    x = "Julian Day",
    y = "Density"
  )

a_hist <- ggplot(depart_flow, aes(x = d_arrive)) +
  geom_histogram(aes(y = ..density..), bins = 52, fill = "grey80", color = "white") +
  geom_density(color = "blue", size = 1) +
  coord_cartesian(xlim = c(1, 365)) +
  geom_vline(xintercept =240) +
  geom_vline(xintercept =136) +
  theme_classic() +
  labs(
    title = "Arrival Timing Distribution",
    subtitle = "Shortnose Sturgeon to Penobscot River",
    x = "Julian Day",
    y = "Density"
  )

png("outputs/arrive_depart_hist.png", width = 2000, height = 1500, res = 300)

(d_hist / a_hist) 

dev.off()

```


## Final Dataset

```{r}
snsdep <- depart_flow %>% 
  filter(!is.na(Mass))
glimpse(snsdep)
```

# Model

## Order categorical variables

```{r}
snsdep$season <- factor(snsdep$season, levels = c("spring", "summer", "fall"))
snsdep$arrive_ssn <- factor(snsdep$arrive_ssn,
                            levels = c("summer", "spring", "fall"))
```

## Set Priors

```{r}
priors <- c(
  # Fixed effects
  set_prior("normal(0, 1)", class = "b", dpar = "musummer"),
  set_prior("normal(0, 1)", class = "b", dpar = "mufall"),
  
  # Intercepts
  set_prior("normal(0, 1)", class = "Intercept", dpar = "musummer"),
  set_prior("normal(0, 1)", class = "Intercept", dpar = "mufall"),
  
  # Random effect SDs
  set_prior("exponential(1)", class = "sd", dpar = "musummer"),
  set_prior("exponential(1)", class = "sd", dpar = "mufall")
)
```

## Models

```{r message=FALSE, warning=FALSE, results='hide'}
set.seed(37)

fit_simp <- brm(
  season ~ arrive_ssn + (1 | animal_id) + (1 | year),
  data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000)

fit_spring_z <- brm(
  season ~ spring_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
  data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000
)

#summary(fit_spring_z)

fit_summer_z <- brm(
  season ~ summer_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
 data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000
)

#summary(fit_summer_z)

fit_fall_z <- brm(
  season ~ fall_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
 data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000
)

#summary(fit_fall_z)

#loo(fit_spring_z, fit_summer_z, fit_fall_z)


fit_spring_summer_z <- brm(
  season ~ spring_flow_z * summer_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
  data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000,
)


fit_spring_fall_z <- brm(
  season ~ spring_flow_z * fall_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
  data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000
)

fit_summer_fall_z <- brm(
  season ~ summer_flow_z * fall_flow_z + arrive_ssn + Mass + day_cap + day_syst +
    (1 | animal_id) + (1 | year),
  data = snsdep,
  family = categorical(),
  prior = priors,   
  chains = 3, cores = 3, iter = 2000, warmup = 1000
)

```

## Compare Models

```{r warning=FALSE, message=FALSE}
loo(fit_simp, fit_spring_z, fit_summer_z, fit_fall_z, fit_spring_summer_z, fit_spring_fall_z,
    fit_summer_fall_z)

final_table
```


### Table 1

```{r warning=FALSE}
loo_simp  <- loo(fit_simp)
loo_spr   <- loo(fit_spring_z)
loo_sum   <- loo(fit_summer_z)
loo_fall  <- loo(fit_fall_z)
loo_ss    <- loo(fit_spring_summer_z)
loo_sf    <- loo(fit_spring_fall_z)
loo_suf   <- loo(fit_summer_fall_z)

pk_summary <- function(loo_obj) {
  tab <- table(cut(loo_obj$diagnostics$pareto_k,
                   breaks = c(-Inf, 0.7, 1, Inf)))
  data.frame(
    good      = tab[1],
    bad       = tab[2],
    very_bad  = tab[3]
  )
}

cmp <- loo_compare(loo_simp, loo_spr, loo_sum, loo_fall,
                   loo_ss, loo_sf, loo_suf)
cmp_df <- as.data.frame(cmp)
cmp_df$model <- rownames(cmp_df)

models <- list(
  fit_simp           = loo_simp,
  fit_spring_z       = loo_spr,
  fit_summer_z       = loo_sum,
  fit_fall_z         = loo_fall,
  fit_spring_summer_z= loo_ss,
  fit_spring_fall_z  = loo_sf,
  fit_summer_fall_z  = loo_suf
)

pareto_df <- do.call(rbind, lapply(models, pk_summary))
pareto_df$model <- rownames(pareto_df)

final_table <- merge(cmp_df, pareto_df, by = "model")
final_table

final_table_3sig <- final_table

num_cols <- sapply(final_table_3sig, is.numeric)

final_table_3sig[num_cols] <- lapply(final_table_3sig[num_cols], signif, digits = 3)

final_table_3sig

write.csv(final_table_3sig, "outputs/model_compare.csv", row.names = FALSE)

```


# Results Figures

## Summary of two best models

```{r}
summary(fit_simp) ## best overall
summary(fit_spring_z) ## best with variables

plot(fit_simp)

```

## Arrival season

```{r}
coefs <- summary(fit_simp)$fixed

# Keep only arrival-season effects
arrive_coefs <- coefs[grep("arrive_ssn", rownames(coefs)), ]
nice_names <- c(
  "musummer_arrive_ssnspring" = "Arrive Spring → Depart Summer",
  "musummer_arrive_ssnfall"   = "Arrive Fall → Depart Summer",
  "mufall_arrive_ssnspring"   = "Arrive Spring → Depart Fall",
  "mufall_arrive_ssnfall"     = "Arrive Fall → Depart Fall"
)

rownames(arrive_coefs) <- nice_names[rownames(arrive_coefs)]



k <- nrow(arrive_coefs)
y <- k:1


```
### Figure 4

```{r}
png("outputs/arrival_effects.png", width = 2000, height = 2000, res = 300)
par(mar = c(6, 15, 4, 2))
plot(arrive_coefs[, "Estimate"], y,
     xlab = "Effect on log-odds (vs spring)",
     ylab = "",
     yaxt = "n",
     pch = 19,
     xlim = range(c(arrive_coefs[, "l-95% CI"], arrive_coefs[, "u-95% CI"])),
     main = "Arrival-season effects on departure timing")

segments(arrive_coefs[, "l-95% CI"], y,
         arrive_coefs[, "u-95% CI"], y)

axis(2, at = y, labels = rownames(arrive_coefs), las = 1)

abline(v = 0, col = "goldenrod", lty = 2)
dev.off()
```


## Flow anomolies

```{r}
coefs <- summary(fit_spring_z)$fixed

flow_coefs <- coefs[grep("spring_flow_z", rownames(coefs)), ]

nice_names <- c(
  "musummer_spring_flow_z" = "Effect of spring flow on Summer vs Spring",
  "mufall_spring_flow_z"   = "Effect of spring flow on Fall vs Spring"
)

rownames(flow_coefs) <- nice_names[rownames(flow_coefs)]



k <- nrow(flow_coefs)
y <- k:1


```

### Figure 5

```{r}
png("outputs/flow_effects.png", width = 2000, height = 1000, res = 300)
par(mar = c(6, 18, 4, 2))
plot(flow_coefs[, "Estimate"], y,
     xlab = "Effect of spring flow anomaly (log-odds)",
     ylab = "",
     yaxt = "n",
     pch = 19,
     xlim = range(c(flow_coefs[, "l-95% CI"], flow_coefs[, "u-95% CI"])))
segments(flow_coefs[, "l-95% CI"], y,
         flow_coefs[, "u-95% CI"], y)
axis(2, at = y, labels = rownames(flow_coefs), las = 1)
abline(v = 0, col = "goldenrod", lty = 2)

dev.off()
```


## Random effects Year & Individual

### Figure 6

```{r}
year_counts <- table(snsdep$year)
id_counts <- table(snsdep$animal_id)

re <- ranef(fit_simp, summary = TRUE)

year_summer <- as.data.frame(re$year[, , "musummer_Intercept"]) %>%
  mutate(year = rownames(re$year))

year_fall <- as.data.frame(re$year[, , "mufall_Intercept"]) %>%
  mutate(year = rownames(re$year))

id_summer <- as.data.frame(re$animal_id[, , "musummer_Intercept"]) %>%
  mutate(animal_id = rownames(re$animal_id))

id_fall <- as.data.frame(re$animal_id[, , "mufall_Intercept"]) %>%
  mutate(animal_id = rownames(re$animal_id))

png("outputs/random_effects.png", width = 2000, height = 2000, res = 300)

par(mfrow = c(2, 2))  

par(mar = c(6, 5, 4, 5))
with(year_summer, {
  plot(Estimate, seq_along(year),
       xlab = "Random intercept (log-odds)",
       ylab = "Year",
       main = "Year effects: Summer vs Spring",
       pch = 19, yaxt = "n")
  axis(2, at = seq_along(year), labels = year, las = 1)
  segments(Q2.5, seq_along(year), Q97.5, seq_along(year))

  axis(4, at = seq_along(year), labels = year_counts[year], las = 1)
  mtext("Number of occurrences", side = 4, line = 3)
})


par(mar = c(6, 5, 4, 5))
with(year_fall, {
  plot(Estimate, seq_along(year),
       xlab = "Random intercept (log-odds)",
       ylab = "Year",
       main = "Year effects: Fall vs Spring",
       pch = 19, yaxt = "n")
  axis(2, at = seq_along(year), labels = year, las = 1)
  segments(Q2.5, seq_along(year), Q97.5, seq_along(year))

  axis(4, at = seq_along(year), labels = year_counts[year], las = 1)
  mtext("Number of occurrences", side = 4, line = 3)
})


par(mar = c(6, 5, 4, 5))
with(id_summer, {
  plot(Estimate, seq_along(animal_id),
       xlab = "Random intercept (log-odds)",
       ylab = "Individual",
       main = "Individual effects: Summer vs Spring",
       pch = 19, yaxt = "n")
  axis(2, at = seq_along(animal_id), las = 1, cex.axis = 0.6)
  segments(Q2.5, seq_along(animal_id), Q97.5, seq_along(animal_id))

  axis(4, at = seq_along(animal_id), labels = id_counts[animal_id], las = 1, cex.axis = 0.6)
  mtext("Number of occurrences", side = 4, line = 3)
})


par(mar = c(6, 5, 4, 5))
with(id_fall, {
  plot(Estimate, seq_along(animal_id),
       xlab = "Random intercept (log-odds)",
       ylab = "Individual",
       main = "Individual effects: Fall vs Spring",
       pch = 19, yaxt = "n")
  axis(2, at = seq_along(animal_id), las = 1, cex.axis = 0.6)
  segments(Q2.5, seq_along(animal_id), Q97.5, seq_along(animal_id))

  axis(4, at = seq_along(animal_id), labels = id_counts[animal_id], las = 1, cex.axis = 0.6)
  mtext("Number of occurrences", side = 4, line = 3)
})

dev.off()

```

